Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-11 paper2.qmd
section:5; process:write

<!-- 3. segway for next section -->
However, because CJ experiments rarely use the exhaustive pairing of sampled judges, stimuli, and individuals--due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020]--a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.

@@

<!-- 2. What is the effect of the comparison mechanism -->
The [DAG -@fig-cj14_dag] also integrates the *comparison mechanism* $C$ into the conceptual-population model, ultimately deriving the outcome $O_{R}$. It shows the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access this outcome due to the comparison mechanism. Moreover, it depicts $C$ as a causal factor influencing $O_{R}$. As with the sampling mechanism, a square encloses $C$ to indicate that it is a conditioned variable--that is, one that determines *which repeated judgments the subset of judges makes for the subset of stimuli produced by the sampled individuals*. In essence, $C$ reflects the assumption that judges *do not* perform all possible repeated judgments for the selected stimuli. Instead, they carry out a "sufficient number of comparisons" $n_{C}$ to allow for accurate estimation of the proportion $P(B>A)$ for each stimulus pair [@Thurstone_1927b, pp. 267].

@@

<!-- 2. assumptions about the outcome, and the sample and comparison mechanism -->
Each CPD in equation ([-@eq-mat51]) rests on specific assumptions, which we outline below. First, we assume that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes. Furthermore, following the conventions of Generalized Linear Latent and Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$. Note that the probability distributions for the sample and comparison mechanisms, $P(S)$ and $P(C)$, included in the probabilistic model [-@fig-cj16_prob], are omitted from the statistical model [-@fig-cj16_stat]. We exclude these probabilities from the statistical definition because, although they are part of the joint distribution of the data, they do not influence the dependence between $O_{R}$ and $D_{R}$, since $S$ and $C$ are governed by random selection designs[^8].

@@

<!-- 2. Continuing with the assumptions -->
Second, we define $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, plus the corresponding repeated judge biases, $B_{JK}[j, k]$. The discriminal processes represent the underlying written-quality traits of the texts. Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as: 
$$
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
$$ {#eq-mat52}
This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022]. In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases). For clarity, the square brackets in $D_{R}$ only indicate the relevant indices within each trait vector; they do not represent any data subsetting.

@@

Third, we define $T_{IA}$ as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables $\beta_{XA}X_{IA}$ (e.g., text length on quality), and the texts' idiosyncratic errors $e_{IA}$. We assume that each element of $e_{IA}$ is independent and identically distributed $(iid)$ according to a Normal distribution, with a mean of zero and a standard deviation $p_{IA}$. The parameter $p_{IA}$ is defined as a proportion of 1 to establish the scale of the latent trait, which is required in any latent variable model [@Depaoli_2021]. Note that because $e_{IA}$ follows a Normal distribution, $T_{IA}$ is also normally distributed. This result indicates that our model reverts to the original assumptions for the discriminal processes of stimuli proposed by @Thurstone_1927b (see @tbl-thurstone_cases).

@@

Sixth, we define $B_{J}$ as a linear combination of the effects of relevant judge-level variables $\beta_{ZJ}Z_{J}$ (e.g., judgment expertise) and judge-specific idiosyncratic errors $e_{J}$. Each element of $e_{J}$ is likewise assumed to be $iid$ Normal with mean zero and a standard deviation $s_{ZJ}$ that depends on the groups to which each judge belongs. For instance, if $Z_{J} = 1, \dots, 3$ represents three groups of judges with varying expertise, the model imposes the constraint $\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1$ to fix the scale of the latent trait [@Depaoli_2021].

>>>>

<!-- 3. segway for next section -->
However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020], CJ experiments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals. Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by certain individuals.

@@

<!-- 2. What is the effect of the comparison mechanism -->
The [DAG -@fig-cj14_dag] also integrates the *comparison mechanism* $C$ into the conceptual-population model, ultimately deriving the outcome $O_{R}$. It shows the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access to $O^{sc}_{R}$ due to the comparison mechanism. Moreover, the DAG depicts $C$ as a causal factor influencing $O_{R}$, enclosing it in a square to indicate that it is a conditioned variable. Specifically, $C$ determines *which repeated judgments the subset of judges makes for the subset of stimuli produced by the sampled individuals*. In essence, $C$ reflects the assumption that judges *do not* perform all possible repeated judgments for the selected stimuli. Instead, they carry out a "sufficient number of comparisons" $n_{C}$ to allow for accurate estimation of the proportion $P(B>A)$ for each stimulus pair [@Thurstone_1927b, pp. 267].

@@

<!-- 2. assumptions about the outcome, and the sample and comparison mechanism -->
Each CPD in equation ([-@eq-mat51]) rests on specific assumptions, which we outline in the [SCM -@fig-cj16_scm], the probabilistic model [-@fig-cj16_scm], and the statistical model [-@fig-cj16_scm]. First, we assume that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes. Furthermore, following the conventions of Generalized Linear Latent and Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$. Note that while the probabilistic model [-@fig-cj16_prob] includes the probability distributions for the sampling and comparison mechanisms, $P(S)$ and $P(C)$, we omit these from the statistical model [-@fig-cj16_stat]. We do so because, although these distributions contribute to the joint distribution of the data, they do not affect the relationship between $O_{R}$ and $D_{R}$. This lack of influence is due to the random selection designs that govern $S$ and $C$[^8].

@@

<!-- 2. Continuing with the assumptions -->
Second, we define $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, plus the corresponding repeated judge biases, $B_{JK}[j, k]$. The discriminal processes represent the underlying written-quality trait of the texts. Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as: 
$$
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
$$ {#eq-mat52}
This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022]. In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases). It is important to clarify that the square brackets in $D_{R}$ merely indicate the relevant indices for each trait vector, without implying any subsetting of the data.

@@

Third, we define $T_{IA}$ as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables $\beta_{XA}X_{IA}$ (e.g., text length on quality), and the texts' idiosyncratic errors $e_{IA}$. We assume that each element of $e_{IA}$ is independent and identically distributed $(iid)$ according to a Normal distribution, with a mean of zero and a standard deviation $p_{IA}$. The parameter $p_{IA}$ is defined as a proportion of 1 to establish the scale of the latent trait, which is required in any latent variable model [@Depaoli_2021]. Note that because $e_{IA}$ follows a Normal distribution, $T_{IA}$ is also normally distributed. This result implies that our model reverts to assuming Normal discriminal processes for the stimuli, as originally proposed by @Thurstone_1927b (see @tbl-thurstone_cases).

@@

Sixth, we define $B_{J}$ as a linear combination of the effects of relevant judge-level variables $\beta_{ZJ}Z_{J}$ (e.g., judgment expertise) and judge-specific idiosyncratic errors $e_{J}$. Each element of $e_{J}$ is likewise assumed to be $iid$ Normal with mean zero and a standard deviation $s_{ZJ}$ that depends on the groups to which each judge belongs. For instance, if $Z_{J}$ is a variable that can take on three values representing three groups of judges with varying expertise, the model imposes the constraint $\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1$ to fix the scale of the latent trait [@Depaoli_2021].
